MASCARA: coexpression analysis in data from designed experiments

### 2024-01-11 R/Paper_markdown.Rmd
24_01_11_2

```{r , echo=FALSE, include=FALSE}
source("DATA_SIM_FUNCS.R")
```

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)
library(MUVR)
library(pheatmap)
library(caret)
library(mixOmics)
library(cowplot)
library(reticulate)
library(biomformat)
library(DiscriMiner)
library(viridis)
library(directlabels)
library(ggpmisc)
library(DescTools)
library(data.table)
library(plotly)
library(ggbreak)
library(rhmmer)
```

@@

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- \bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{T}_{\beta}\bm{P}_{\beta}^T + \bm{T}_{S}\bm{P}_{S}^T + \bm{E} \\ -->
<!-- \text{where for each element}\\ -->
<!-- &\bm{T}_{S_{ik}} \sim \mathcal{N}(0,1), \quad \bm{P}_{S_{jk}}^{\text{baits,spikes}} \sim \mathcal{N}(0.5,0.2) \\ -->
<!-- \text{and:} &\\ -->
<!-- \bm{P}_{S}^{\text{rest}} &= 0 -->
<!-- \end{aligned}(\#eq:Sim2-bp) -->
<!-- \end{equation} -->

<!-- The expression levels of differential transcripts is modeled with the scores and loadings of the growth condition and its interaction with time; termed here the combined effect ( $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{P}_{\alpha + \alpha\beta}^T$ ). The time effect is modeled with scores and loadings ( $\mathbf{T}_{\beta}$  and $\mathbf{P}_{\beta}^T$ ). Also included is some random measurement noise, this is modeled as a matrix $\mathbf{E}$ with normally distributed random values $e_{ij} = \mathcal{N}(0,\sigma^2)$.  -->

<!-- On top of the between-group effect there is an additional correlation structure applied to the baits and spikes to create a model of coexpression that is independent to the variance induced by the experimental factors ($\mathbf{X}_S = \mathbf{T}_S\mathbf{P}_S^T$). This $\mathbf{X}_S$ matrix contains two component sources of variation which links the baits and spikes. In this simulation we have a set of 4 baits that have positive loadings for the combined effect along with a variable number of other genes - "Combined Effect Responders", which are the genes from a simulated URP (indicated by the yellow bar in Figure \@ref(fig:simOverview)). For the structured ambient variance; 2 random vectors in \(\bm{T}_S\) (each column representing an effect caused by ambient conditions) which our baits and spikes contribute to (i.e. green and black bars in Figure \@ref(fig:simOverview) are non zero in \(\bm{P}_S\)). Similarly to the real data, the gene expression values for each experimental condition are simulated with three independent replicates per time-condition-combination. Each simulated data $\mathbf{Y}$ set thus consists of 24 samples and 2000 genes. -->

@@

<!-- The scores matrices $\mathbf{T}_{\alpha + \alpha\beta}$ and $\mathbf{T}_{\beta}$ each consist of two orthogonal score vectors and their relative levels are indicated by the colour. The different colours in Figure \@ref(fig:simOverview) represent that the scores are orthogonal to each other. Each small block consists of the same score value for each of the replicates within that condition. Loadings $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$  are simulated to create the baits and spikes. As displayed in Figure \@ref(fig:simOverview), the green bar represents 4 baits (or known pathway genes), the black bar represents the 12 spikes (“unknown” genes from POI or RP), yellow represents differential URP genes, and purple; a set of differential URPs with an alternative treatment response.  -->

<!-- In $\mathbf{P}_{\beta1}^T$ the baits, spikes and other DE genes (all genes from POI, RP and URP) along with the remaining first half have similar positive values (represented by light grey) where the other 1000 have positive values in $\mathbf{P}_{\beta2}^T$ (grey), these are created to emulate the reality of the real data in which many genes are slightly correlated with eachother through the course of development, as well as to account for some nonlinearities which are otherwise ignored by our model. The remaining elements of $\mathbf{P}_{\alpha + \alpha\beta}^T$ and $\mathbf{P}_{\beta}^T$ are populated with zeroes. The $\mathbf{T}$ matrices represent the simulated scores for their respective factors. For blues, reds, greys and golds we have 4 different shades representing 4 unique values per factor level (total 8 unique values where reds and golds are sign flipped blues and greys respectively). Similarly oranges and pinks represent a growth condition independent time profile with four unique values. $\mathbf{E}$ represents random unstructured technical noise and the additional $\bm{1\mu}^T$ term in \@ref(eq:Sim2-bp) represents a mean value for each gene. -->

<!-- The issue to address is the presence of many differential URPs; those that have strong between group correlations but no within group correlation. Here these differential genes relate  to the yellow bar in Figure \@ref(fig:simOverview). We incrementally increase the number of non-zero values and decrease the number of 0 values, which in figure \@ref(fig:simOverview) relates to increasing the size of the yellow bar, as well as varying the size (as quantified by the Frobenius norm) of the combined effect matrix through the scalar parameter *d* below; -->

@@

<!-- In order to evaluate the methods and indicate the levels of variance required for optimal performance we conduct a larger scale assay. Here we fix the number of DE URP genes to 30 which corresponds to the yellow bar size and the size of the within correlation effect matrix (\(\|\bm{T}_S\bm{P}_S^T\|_F = 0.025\)). We investigate a range of values for sizes of between variance (*d*) and random noise (*l*) across random structures (which correspond to repeated simulations of randomised values for all $\mathbf{T}$ matrices). Parameters were adjusted according to \@ref(eq:Test-2) on 20 randomised $\mathbf{T}$ matrices. -->

@@

   <!-- c. Perform linear regression of \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) onto their corresponding scores \( \bm{t}_k \) and \( \bm{w}_k \) to obtain the respective loadings \( \bm{p}_k \), and \( \bm{q}_k \). -->
   <!-- d. Deflate \( \mathbf{\tilde{Y}}_k \) and \( \mathbf{B}_k \) by removing the variance explained by the $k^{th}$ set of scores in the space of $\mathbf{\tilde{Y}}$\( \bm{t}_k \) and use the deflated matrices as inputs for the next iteration. -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- \mathbf{\tilde{Y}_{k}} = \mathbf{\tilde{Y}}_{k-1} - \mathbf{t}_{k}\mathbf{p}^T_{k} \\ -->
<!-- \text{and} \\ -->
<!-- \mathbf{B_{k}} = \mathbf{B}_{k-1} - \mathbf{t}_{k}\mathbf{q}^T_{k} \\ -->
<!-- \end{aligned} -->
<!-- \end{equation} -->

@@

```{r, include = F}
source("DATA_SIM_FUNCS.R")

@@

```{r}
all <- cbind.data.frame(meta[,c("growth_condition","time")], X)
residuals <- cbind.data.frame(meta[,c("growth_condition","time")], res_ASCAplus_nb$residuals)
```

```{r}
#get cands from residuals
rPCD <- prcomp(data.matrix(residuals[,-c(1:2)]))
```

```{r}
cutoff <- round(quantile(c(1:nrow(ASCA_cands)),0.1))
res_filt <- residuals[,c(1:2,which(colnames(residuals) %in% rownames(ASCA_cands)[1:cutoff]))]

rpcd <- prcomp(data.matrix(res_filt[,-c(1:2)]))

ref$Baits <- as.character(ref$Baits)
ref$Baits[which(ref$Feature %in% PSI_ref[,1])] <- "PSI"
ref$Baits <- factor(ref$Baits, levels = c("Other","Pathway","PSI"))

Fref <- ref[match(rownames(rpcd$rotation),ref$Feature),]

```{r, include = F}
importance <- paste0(round(summary(rpcd)$importance[2,] * 100, digits = 2), "%")
```

```{r}
# resids <- residuals[,-c(1:2)]
resids <- res_filt[,-c(1:2)]
spls_res <- spls(resids[,-which(colnames(resids) %in% baits)], resids[,which(colnames(resids) %in% baits)], all.outputs = T)

sPLS_cands <- cbind.data.frame(abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X, spls_res$loadings$X * spls_res$prop_expl_var$X)
sPLS_cands <- sPLS_cands[order(sPLS_cands[,1], decreasing = T),]

importance <- paste0(round(spls_res$prop_expl_var$X * 100, digits = 2), "%")

Fref2 <- Fref[-which(Fref$Feature %in% baits),]
Fref2 <- Fref2[match(rownames(sPLS_cands),Fref2$Feature),]

```{r}
#calculating direction of baits to determine direction of candidate selection spls loading space

angles <- rbind.data.frame(spls_res$loadings$Y, colMeans(spls_res$loadings$Y))

x <- angles[4,]
y <- angles[5,]

#adapted from https://stackoverflow.com/questions/1897704/angle-between-two-vectors-in-r
angle <- function(x,y){

  x <- c(as.matrix(x))
  y <- c(as.matrix(y))

  dot.prod <- x%*%y 
  norm.x <- norm(x,type="2")
  norm.y <- norm(y,type="2")
  theta <- acos(dot.prod / (norm.x * norm.y))
  return(as.numeric(theta))
}

angles_loadings <- apply(sPLS_cands[,-1],1,angle, y = y)

#define this better
sPLS_cands <- cbind.data.frame(sPLS_cands, "angle" = angles_loadings, "score" = 3 * sPLS_cands$`abs(spls_res$loadings$X) %*% spls_res$prop_expl_var$X` * (1-angles_loadings))

sPLS_cands <- sPLS_cands[order(sPLS_cands$score, decreasing = T),]
```

```{r}
avg_point <- angles[5,]/5

rot.mat <- matrix(c(0,-1,1,0),nrow = 2, byrow = T)  #90 degree clockwise rotation

rotated_point <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat)

ang <- angle(avg_point,c(1,0))

rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2)

```{r}
#rotated loadings
ang <- angle(avg_point,c(1,0))

rot.mat2 <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation

rotated_point2 <- as.data.frame(c(as.matrix(avg_point)) %*% rot.mat2)

rotate <- function(x,y){
  x <- c(as.matrix(x))
  x %*% y
}

loadings_r_baits <- t(apply(sPLS_cands[,c("comp1","comp2")],1,rotate, y = rot.mat2))
colnames(loadings_r_baits) <- c("comp1","comp2")
```

```{r}
sPLS_bait_cands <- loadings_r_baits[order(loadings_r_baits[,1], decreasing = T),]

```{r}
range.ang <- seq(from = -pi/6, to = pi/6, length.out = 60)

###

i <- NULL
L_R <- data.frame()
LOADINGS_R <- list()

for(i in 1:length(range.ang)){

ang <- range.ang[i]

rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
loadings_r <- t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat))
colnames(loadings_r) <- c("comp1","comp2")
L_R <- rbind(L_R,loadings_r)
  LOADINGS_R[[i]] <- loadings_r

# colnames(LOADINGS_R) <- c("comp1","comp2")
L_R$rad <- rep(range.ang, each = nrow(sPLS_cands))
L_R$frame <- rep(c(1:length(range.ang)), each = nrow(sPLS_cands))

```{r}
#get sd of original comp1 distribution
sigma <- sd(sPLS_cands$comp1)

big.T.sig <- function(x,y){
  big <- length(which(x > 2.5*y))
  return(big)
}

i <- NULL
res <- c()
for(i in 1:length(LOADINGS_R)){

  r <- big.T.sig(LOADINGS_R[[i]][,1],sigma)
  res <- c(res,r)
}

```{r}
RES <- cbind.data.frame("Rad" = range.ang, "Density" = res)

density <- ggplot(RES, aes(x = Rad, y = Density))+
  geom_smooth() + theme_bw() + ggtitle("Number of points above 2*sigma of original rotation")

#taken from https://stackoverflow.com/questions/6836409/finding-local-maxima-and-minima
inflect <- function(x, threshold = 1){
  up   <- sapply(1:threshold, function(n) c(x[-(seq(n))], rep(NA, n)))
  down <-  sapply(-1:-threshold, function(n) c(rep(NA,abs(n)), x[-seq(length(x), length(x) - abs(n) + 1)]))
  a    <- cbind(x,up,down)
  list(minima = which(apply(a, 1, min) == a[,1]), maxima = which(apply(a, 1, max) == a[,1]))
}

ggres <- ggplot_build(density)$data[[1]]
infs <- inflect(ggres$y)
rads <- ggres$x[infs$maxima]

```{r}
maxima.ang <- rads

###

i <- NULL
sPLS_maxima_cands <- list()

for(i in 1:length(maxima.ang)){

ang <- maxima.ang[i]

rot.mat <- matrix(c(cos(ang),-sin(ang),sin(ang),cos(ang)),nrow = 2, byrow = T) #clockwise rotation
loadings_r <- as.data.frame(t(apply(loadings_r_baits[,c("comp1","comp2")],1,rotate, y = rot.mat)))
colnames(loadings_r) <- c("comp1","comp2")
sPLS_maxima_cands[[i]] <- loadings_r[order(loadings_r$comp1, decreasing = T),]

```{r}
Fref2 <- Fref[-which(Fref$Feature %in% baits),]
Fref2 <- Fref2[match(rownames(sPLS_maxima_cands[[1]]),Fref2$Feature),]
```

```{r}
Fref2 <- Fref[-which(Fref$Feature %in% baits),]
Fref2 <- Fref2[match(rownames(sPLS_maxima_cands[[2]]),Fref2$Feature),]

@@

residual_cands <- ranked_dist(baits[1:3], rpcd)

residual_coexp <- ranked_coexp(baits, resids)

sPLS_cands <- sPLS_maxima_cands[[2]]

Methods <- c("sPLS_ASCA","PCA_DIST_ASCA","Correlations_ASCA", "Correlations")

getrank <- function(RES,geneset,NOINFO = 26402){

  ranks <- which(rownames(RES) %in% geneset)
  if(length(ranks) < length(geneset)){
    ranks <- c(ranks,rep(nrow(RES) + 1, length(geneset) - length(ranks))) #NOINFO
  }

  Ranks <- cbind.data.frame("Geneset" = geneset, "Rank" = ranks)
  return(Ranks)

psis <- PSI_ref[which(PSI_ref[,1] %in% colnames(resids)),1]

```{r}
GA_ref <- read_xlsx("Rice_Annotation/GA pathway genes-IDs.xlsx")
GA_ref$ID <- gsub("g","t",GA_ref$ID)
GAs <- ref[which(gsub("-.*","",ref$Feature) %in% GA_ref$ID),1]

GAs <- GAs[which(GAs %in% colnames(resids))]
sPLS_ASCA <- rbind.data.frame(getrank(sPLS_cands,p_n_b),getrank(sPLS_cands,psis),getrank(sPLS_cands,GAs))
sPLS_ASCA <- cbind.data.frame(sPLS_ASCA, "Method" = rep("sPLS_ASCA", nrow(sPLS_ASCA)))

#####

PCA_DIST_ASCA <- rbind.data.frame(getrank(residual_cands,p_n_b),getrank(residual_cands,psis),getrank(residual_cands,GAs))
PCA_DIST_ASCA <- cbind.data.frame(PCA_DIST_ASCA, "Method" = rep("PCA_DIST_ASCA", nrow(PCA_DIST_ASCA)))

Correlations <- rbind.data.frame(getrank(coexp_cands,p_n_b),getrank(coexp_cands,psis),getrank(coexp_cands,GAs))
Correlations <- cbind.data.frame(Correlations, "Method" = rep("Correlations", nrow(Correlations)))

Correlations_ASCA <- rbind.data.frame(getrank(residual_coexp,p_n_b),getrank(residual_coexp,psis),getrank(residual_coexp,GAs))
Correlations_ASCA <- cbind.data.frame(Correlations_ASCA, "Method" = rep("Correlations_ASCA", nrow(Correlations_ASCA)))

RES <- rbind.data.frame(sPLS_ASCA, PCA_DIST_ASCA, Correlations_ASCA, Correlations)

RES$Geneset[which(RES$Geneset %in% p_n_b)] <- "SLs"
RES$Geneset[which(RES$Geneset %in% psis)] <- "PSIs"
RES$Geneset[which(RES$Geneset %in% GAs)] <- "GAs"

RES$Method[which(RES$Method == "sPLS_ASCA")] <- "MASCARA"

RES <- RES[which(RES$Method %in% c("Correlations","MASCARA")),]

q <- ggplot(RES[-which(RES$Geneset == "GAs"),], aes(x = Method, y = Rank, fill = Geneset)) + geom_boxplot() + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust=1))
```

```{r, fig.cap="Method Ranks on Real Data"}

REF2 <- as.data.frame(read_tblout("Rice_Annotation/HMM_ITAG_1.txt"))

f_p_m <- REF2[which(REF2$domain_name %in% rownames(sPLS_cands)[1:25]),]
f_p_m <- f_p_m[order(f_p_m$sequence_evalue),]
fpm <- f_p_m[!duplicated(f_p_m$domain_name),]
fpm <- fpm[,c(1,3,19)]
fpm[1,3] <- "Diterpenoid phytoalexin biosynthesis"
fpm[12,3] <- "Vesicle-Associated Membrane Protein-Associated Protein"
cands <- rownames(sPLS_cands)[1:25]
cands <- cands[which(cands %in% fpm$domain_name)]
rownames(fpm) <- fpm[,1]
fpm <- fpm[cands,]

```{r}
rownames(fpm) <- NULL

@@

```{r realcandidates, fig.cap="Selected candidates from A-C ranked correlations and D-F MASCARA. X axes; transformed expression level of one of the bait genes (CCD8), y axes; expression levels of three example coexpression candidates for each method (see @haiderTranscriptomeAnalysisPhosphate2023 for preprocessing details)."}

qqplots_spls <- list()
qqplots_coexp <- list()
qqplots_pca <- list()
qqplots_resid_cor <- list()

qqplots_spls_SR <- list()

i <- NULL
for(i in 1:9){
  qqplots_spls[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(sPLS_cands)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(sPLS_cands)[i])

@@

  qqplots_resid_cor[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(residual_coexp)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(residual_coexp)[i])

@@

plot1 <- wrap_plots(qqplots_coexp[c(5,7,8)]) + plot_annotation(title = "Correlations")
plot2 <- wrap_plots(qqplots_spls[c(1,5,8)]) + plot_annotation(title = "MASCARA")

# plot1/plot2 + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1'))

@@

We acknowledge funding by the Dutch Research Council (NWO/OCW) for the MiCRop Consortium program, Harnessing the second genome of plants (Grant number 024.004.014; to HB, LD, AKS, JAW), the Dutch Research Council (NWO-TTW grant 16873 Holland Innovative Potato; to HB and LD) and the Data Science Centre of the University of Amsterdam (to FW). FW would like to thank Frans van der Kloet, Fentaw Abegaz, Roel van der Ploeg for statistical discussions and Imran Haider for consultation on the strigolactone pathway.

@@

<!-- $r_{j_{sl}j_{q}} = \frac{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl}) (y_{ij_{q}} - \bar{y}_j_{q})}{\sqrt{\sum_{i=1}^{I} (y_{ij_{sl}} - \bar{y}_j_{sl})^2 \sum_{i=1}^{I} (y_{ij_{q}} - \bar{y}_j_{q})^2}}$ -->

<!-- MASCARA example run on $params: -->

<!-- Data was simulated: -->

<!-- \begin{equation} -->
<!-- \begin{aligned} -->
<!-- \bm{Y} &= \bm{1\mu}^T + \bm{T}_{\alpha + \alpha\beta}\bm{P}_{\alpha + \alpha\beta}^T + \bm{Y}_{\beta} + \bm{E} \\ -->
<!-- \text{where:} &\\ -->
<!-- \bm{T}_{\alpha + \alpha\beta} &= \text{vec}(\bm{ab}^T) \quad \text{where:} \\ -->
<!-- \bm{c}_{i,1:4} &\sim \mathcal{N}(0,1), \quad \bm{a}_{1:r} = [1,\ldots,1], \quad \bm{b}_{1:8} = [\bm{c}_{i},-\bm{c}_{i}]^T \\ -->
<!-- \bm{T}_{\beta} &= \text{vec}(\bm{de}^T) \quad \text{where:} \\ -->
<!-- \bm{f}_{1:4} &\sim \mathcal{N}(0,1.5), \quad \bm{d}_{1:r} = [1,\ldots,1], \quad \bm{e}_{1:8} = [\bm{f},\bm{f}]^T \\ -->
<!-- \bm{P}_{\alpha + \alpha\beta}^{1:2} &= \begin{bmatrix}1.5_{1:4} & 1_{5:8} & 0.5_{9:12} & 0_{13:2000}\end{bmatrix} \\ -->
<!-- \bm{P}_{\beta}^{1:2} &= \begin{bmatrix}0.1_{1:1000} & 0_{1001:2000}\end{bmatrix} \\ -->
<!-- \bm{E} &\sim \mathcal{N}(0,0.75) -->
<!-- \end{aligned}(\#eq:suppl-Sim-bp) -->
<!-- \end{equation} -->

<!-- genes 1:12 have non-zero loadings on PC1 of $svd({\bm{Y}_{\alpha + \alpha\beta}})$, these are considered genes of interest. These 12 genes correspond to the baits and spikes (spikes represent "unknown" genes of interest, used for method evaluation). Which of these genes are baits and which are spikes depends on the methods being tested and will be discussed in the Analysis and Feature Selection section. genes 501:510 have a value of 0.5 on PC2 of $svd({\bm{Y}_{\alpha + \alpha\beta}})$ and are considered to display a nonsense effect. -->

<!-- genes 1:1000 have a loadings of 0.1 on PC1 of $svd({\bm{Y}_\beta})$ and genes 1001:2000 have loadings of 0.1 on PC2 of $svd({\bm{Y}_\beta})$. To create $\bm{Y}$ the dot product of the scores and transposed loadings is calculated per factor, and added to a general mean value for each feature ($\bm{1\mu}^T$) as well as some normally distributed noise \(\bm{E}\) to create differences between the replicates, see equation \@ref(eq:Sim-bp) for an overview. -->

<!-- <!-- \End{multicols} -->

<!-- sim2:  -->
<!-- For this simulation we create 2 components in the combined effect (${\bm{Y}_{\alpha + \alpha\beta}}$), 2 for time ($\bm{Y}_\beta$). For the time effect, we create this similarly to simulation 1. For the combined effect: 2 random orthogonal scores. The loadings for these components contain several sets of genes, some with positive and some with negative loadings onto the 4 components.  -->

<!-- Y = TP (ambient structure) This is regressed onto the concatenated scores of the combined effect and time to ensure orthogonality, creating relationships that are independent of the experimentally induced variance. -->

>>>>

```{r , echo=FALSE, include=FALSE}
source("DATA_SIM_FUNCS.R")
source("MASCARA_FUNCS.R")
```

```{r, echo = FALSE, include=FALSE}
library(devtools)
library(ggplot2)
library(MetStaT)
library(gASCA)
library(patchwork)
library(readxl)
library(ggfortify)
library(ggrepel)
library(grid)
library(gridExtra)
library(tidyr)
library(tidyverse)
library(reshape2)
library(scales)
library(matrixStats)
library(DESeq2)
library(MASS)
library(glmnet)
library(doParallel)
library(MUVR)
library(pheatmap)
library(caret)
library(mixOmics)
library(cowplot)
library(reticulate)
library(biomformat)
library(DiscriMiner)
library(viridis)
library(directlabels)
library(ggpmisc)
library(DescTools)
library(data.table)
library(plotly)
library(ggbreak)
library(rhmmer)
library(pracma)
```

@@

@@

@@

@@

@@

```{r, include = F}
source("DATA_SIM_FUNCS.R")
source("MASCARA_FUNCS.R")

@@

```{r}
all <- cbind.data.frame(meta[,c("growth_condition","time")], X)
residuals <- cbind.data.frame(meta[,c("growth_condition","time")], res_ASCAplus_nb$residuals)

cutoff <- round(quantile(c(1:nrow(ASCA_cands)),0.1))
res_filt <- residuals[,c(1:2,which(colnames(residuals) %in% rownames(ASCA_cands)[1:cutoff]))]

resids <- res_filt[,-c(1:2)]

@@

sPLS_cands <-  MASCARA(data.matrix(resids), baits)[[1]]             #MASCARA

@@

```{r realcandidates, fig.cap="Example candidates from A-C ranked correlations and D-F MASCARA. X axes; transformed expression level of one of the bait genes (CCD8), y axes; expression levels of three example coexpression candidates for each method (see @haiderTranscriptomeAnalysisPhosphate2023 for preprocessing details)."}

qqplots_spls <- list()
qqplots_coexp <- list()

i <- NULL
for(i in 1:20){
  qqplots_spls[[i]] <- qqplot(cbind.data.frame("Candidate" = X[,which(colnames(X) %in% rownames(sPLS_cands)[i])],
                        "Bait" = X[,which(colnames(X) %in% baits[1])]),meta, 
       aes(x = Bait, y = Candidate, colour = growth_condition, shape = time)) + ylab(rownames(sPLS_cands)[i])

@@

@@

plot1 <- wrap_plots(qqplots_coexp[c(6,8,10)]) + plot_annotation(title = "Correlations") #+ theme(aspect.ratio = 1)
plot2 <- wrap_plots(qqplots_spls[c(2:4)]) + plot_annotation(title = "MASCARA") #+ theme(aspect.ratio = 1)

# wrap_plots(qqplots_spls) + plot_layout(guides = "collect",tag_level = 'new')

PLOT <- plot1/plot2 + plot_layout(guides = "collect",tag_level = 'new') + plot_annotation(tag_levels = c('A','1'))
# ggsave("within_between_example.png", PLOT)

@@

We acknowledge funding by the Dutch Research Council (NWO/OCW) for the MiCRop Consortium program, Harnessing the second genome of plants (Grant number 024.004.014; to HB, LD, AKS, JAW), the Dutch Research Council (NWO-TTW) Holland Innovative Potato (grant 16873; to HB, LD, FW) and the Data Science Centre of the University of Amsterdam (to FW). FW would like to thank Frans van der Kloet, Fentaw Abegaz, Roel van der Ploeg for statistical discussions and Imran Haider for consultation on the strigolactone pathway.

@@
